---
title: "class12 Homework"
author: "Audrey Ting Zhu (A16898668)"
date: "2024-11-11"
output: html_document
---


Section 4: Population Scale Analysis [HOMEWORK]

>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr<-read.table("rs.txt")
head(expr)
```
>Number of samples: 

```{r}
nrow(expr)
```
>sample size for each genotype: 

```{r}
table(expr$geno)
```

> Expression levels medians(looked at stackoverflow): 

```{r}
library(dplyr)
```

```{r}
median_genotype_exp <- expr %>%
  group_by(geno) %>%
  summarize(Median_express = median(exp,na.rm = TRUE))

median_exp<- table(median_genotype_exp$Median_express)

print(median_exp)
```
respectively for G/G A/G A/A .


>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?


>ANS: The SNP affects the expression. Clearly the G/G genotype leads to reduced gene expression compared to the A/A genotype. The A/G genotype leads to an expression level between the two. This shows how the having the G SNP leads to reduction in expression. 


```{r}
library(ggplot2)

```
Lets make a boxplot

```{r}
ggplot(expr)+aes(x=geno,y=exp,fill=geno)+geom_boxplot(notch=TRUE)
```

