---
title: "Class 06 R functions"
author: "Audrey Ting Zhu (A16898668)"
date: "2024-10-21"
output: pdf_document
---


This week we are introducing R functions and how to write our own
Questions to answer:


Q1.  Write a function grade() to determine an overall grade from a vector of student homework
 assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
 NA value) this can be used as a score to be potentially dropped. 
 
```{r}
# Example input vectors to start with
 
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```
 
```{r}
#finding the element showing lowest grade
which.min(student1)
```
 I now need to exclude the lowest element
```{r}
student1[-8]
#This excludes the lowest value manually. Replace 8 with which.min(student1)
student1[-which.min(student1)]
```

Now we can find mean:
```{r}
mean(student1[-which.min(student1)])
```
However, this wouldn't work on the other vectors in students, as they contain NA. We need to replace NA values with 0. First, we need to find the NA elements.

```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
x<-student2
is.na(x)
which(is.na(x))
```
Now we replace NA with 0
```{r}
x[is.na(x)]<-0
x
```
Put it all together
```{r}
#for student2
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
x<-student2
x[is.na(x)]<-0
mean(x[-which.min(x)])

#for student3
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x<-student3
x[is.na(x)]<-0
mean(x[-which.min(x)])
```
Now make it a function.
A function has 3 parts:
1. A name (ex.grade())
2. input arguments(vector of student scores)
3. The body, working snippet of code
```{r}
grade <- function(x) {
  x[is.na(x)]<-0
  mean(x[-which.min(x)])
}
```

Check function
```{r}
grade(student1)
grade(student2)
grade(student3)
```

Adding comments so others can use the function
```{r}

#' Calclates the average score for a vector of student scores, lowest grade dropped, missing grades are counted as 0
#'
#' @param x Numeric value of student scores
#'
#' @return Average score
#' @export
#'
#' @examples student<-c(90,97,90,NA)
#'              grade(student)

grade <- function(x) {
  #missing grades(NA)counted as zero because they are missing
  x[is.na(x)]<-0
  #lowest score is excluded before mean calculation
  mean(x[-which.min(x)])
}

```

Now we use our function on real whole class data.
Class data is from this CSV format:  “https://tinyurl.com/gradeinput”
 
```{r}
url<-"https://tinyurl.com/gradeinput"
gradebook<-read.csv(url,row.names=1)
```
 
```{r}
apply(gradebook,1,grade)
```
 
 Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student
 overall in the gradebook? 
 
 We first save the grades calculated from the csv file into results.
 
```{r}
results <-apply(gradebook,1,grade)

```
 Now I find the top scoring 
 
 
```{r}
which.max(results)
```
 Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained
 the lowest scores overall? 
 
 
```{r}
gradebook
```

 Now we are looking at the averages of columns (the specific hw). Thus, the margin will be 2, instead of 1. 
 
```{r}
#find the average scores of HW, NA must be set to TRUE, in order for function to calculate.
ave.scores<-apply(gradebook,2,mean, na.rm=TRUE)
ave.scores
#Finding the worst HW by searching for the lowerst average score
which.min(ave.scores)
```
 The worst HW seems to be 3. 
 
However, sometimes mean is not the best indicator, as it is sensitive to outliers. We will use median.
 
```{r}
median.scores<-apply(gradebook,2,median, na.rm=TRUE)
median.scores
which.min(median.scores)
```
 In this case, the worst HW is 2. 
 
 So which is the correct answer? We can use graphs to give a clearer answer by visualizing distribution. 
 
 
```{r}
boxplot(gradebook)
```
 
 There is a really great distribution for HW2. Some did well. Some did horrible. HW3 does not have this wide distribution. 
 
 
 
  Q4. Optional Extension: From your analysis of the gradebook, which homework was most
 predictive of overall score (i.e. highest correlation with average grade score)?
 
 We hope that the students average score(grade) correlates to the scores they get on the hw (gradebook columns). 
 
```{r}

#mask the NA scores in the gradebook 
masked.gradebook<-gradebook
masked.gradebook[is.na(masked.gradebook)]<-0
masked.gradebook
```

Find correlation function
```{r}
cor(results, masked.gradebook$hw5)
```
 
Apply to all gradebook
```{r}
apply(masked.gradebook,2,cor,x=results)
```
It seems hw 5 has the highest correlation. Hw5 is most predictive. 



 Q5. Make sure you save your Quarto document and can click the “Render” (or Rmark
down”Knit”) button to generate a PDF foramt report without errors. Finally, submit your PDF
 to gradescope. 
 