---
title: "Halloween Mini Project"
author: "Audrey Ting Zhu (A16898668)"
date: "2024-11-03"
output: html_document
---
##Class 10: Halloween Mini-Project


```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file,row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?
>ANS: There are 85 candy types.

```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?
>ANS: There are 38 fruity candy types. 

```{r}
sum(candy$fruity==1)
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
> ANS: My favorite candy is Air Heads. The winpercentage is 52.34146.

```{r}
candy["Air Heads", ]$winpercent
```


>Q4. What is the winpercent value for “Kit Kat”?
>ANS:76.7686

```{r}
candy["Kit Kat", ]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
>ANS:49.6535

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent

```
```{r}

library("skimr")
skim(candy)
```


>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

>ANS: Winpercent is in a different range as it is a percentage score ranging up to 100. The other variables scale up to 1. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

>ANS: 0 means that there is no chocolate in the candy composition. 1 means that there is chocolate in the candy.  

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy, aes(x=winpercent))+geom_histogram(binwidth=5)+labs(title="Winpercentages for Candies", x="Winpercent")
```



>Q9. Is the distribution of winpercent values symmetrical?

>ANS: The distribution of winpercent values are not 100 percent symmetrical, but it is roughly. 

>Q10. Is the center of the distribution above or below 50%?

>ANS: Depends if you look at mean or median. The mean is slightly above 50 but the median is 47.82, which is below 50%. 

```{r}
mean(candy$winpercent)
median(candy$winpercent)
```

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

>ANS: Chocolate candy has an higher score. 

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```

>Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)],candy$winpercent[as.logical(candy$fruity)])
```
>ANS: P-value is less than 0.05. It is pretty significant. 


>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(-candy$winpercent),], n=5)
```

>Q14. What are the top 5 all time favorite candy types out of this set?
ANS: The benefit of dplyr is that it has much simpler and readable syntax. However, you also need to download a package to make it work. 

```{r}

library("dplyr")
```
```{r}

candy %>% arrange(winpercent) %>% head(5)
```


>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}

library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_bar(stat="identity")
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_bar(stat="identity")
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
>Q17. What is the worst ranked chocolate candy?
>ANS: Apparently people dont sixlets.
>Q18. What is the best ranked fruity candy?
>ANS: People like Starburst. 


```{r}
library("ggrepel")
```

```{r}


ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
  
```
>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
ANS: Reeses Minatures



>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

>ANS: Nik L Lip manages to be expensive and unliked the most. 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()


```

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

>ANS: Chocolate and fruity are anti-corellated, which is a shame because I like chocolate fruity candies. 

>Q23. Similarly, what two variables are most positively correlated?
>Ans: Chocolate and winpercent are correlated. Chocolate and bar are also correlated. 






```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
pca$rotation[,1]
```

```{r}

plot(pca$x[,1:2],col=my_cols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}

library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
>ANS:Fruity, hard and pluribus are heavily picked up by PC1. This makes sense. Afterall, many candies have the characteristics of being packed with multiple pieces in a bag, being fruity, and being hard overlap together. 